Exploring the Presidents Corpus

### 2019-10-17 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
update on Presidents paper

\usepackage{arxiv}
\usepackage{float}

@@

Alzheimer's Disease (AD) and other forms of dementia affect a significant proportion of the older age population in the world today and is currently the sixth leading cause of death in the US and was named the leading killer of women in the UK. According to a recent report commissioned by the Alzheimer's Society in 2015, they estimate the prevalence of AD in the UK at approximately 815,000 people. This represents 1 in 14 of those aged 65 or over and 1 in 79 of the general population \cite{AlzheimersSociety2014}. From a financial perspective, they estimate an annual spend of £4.3 billion of which approximately £85 million is spent solely on diagnosis and that the total impact of AD (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars.

@@

\subsubsection{Linguistic Inquiry and Word Count(LIWC)}
The LIWC is a text processor which analysis words within a given a document and compares the words in the LIWC dictionary file. The LIWC dictionary file contains over 6000 words which are categorised into approximately 90 different types. For example the word 'cried' is part of five word categories: sadness, negative emotion, overall affect, verbs and past focus \cite{Pennebaker2015}. Each word in the document or set of documents is searched for in the dictionary file and if found, the appropriate type is incremented by 1. What is output is an analysis of word usage in reference to each category. This is particularly useful in analysing both structural language changes but also the content or themes of language for a given transcript or set of transcripts.
\par 
As with Berisha and Liss, we analysed the transcripts over time in order to determine if there was an underlying temporal trend that may indicate that different parts of language increase or decrease over time. In AD, cognitive abilities are said to deteriorate over time and therefore it may be useful to analyse temporal trends that indicate a potential deterioration of cognition.

\section{Results}\label{results}
One of the most important thing to note is the wide variety of samples between the three presidents and also the varying timescales. RR participated in 46 press conferences over eight years (an average of 5.75 a year) which is the fewest number of press conferences given by an American president during their term of office. GHWB participated in 136 press conferences over four years (an average of 34 a year) and DJT participated in 29 press conferences to date (an average of 19.3 per year). Equally, there are differences in the average number of words. RR produced an average of 3424 words per conference compared to 2608 by GHWB (unpaired t = 4.434, p\textless 0.001) and DJT at 1849 words (unpaired t = 6.524, p\textless 0.001).
\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} |}
		\hline
		& RR & GHWB & DJT \\ \hline
		Total Words & 3423.91 (416.42) & 2607.72 (1210.38) & 1848.65 (1549.38) \\ \hline
		Unique Words & 894.13 (85.15) & 667.76 (218.67) & 481.82 (221.29) \\ \hline
		Mean Length of Utterance & 23.17 (1.402) & 18.71 (2.067) & 481.82 (221.29) \\ \hline
		Non Specific Nouns & 12.72 (4.63) & 6.78 (4.32) & 7.41 (8.75) \\ \hline
		LI Verbs & 124.22 (17.89)& 103.45 (51.87) & 84.48 (75.78) \\ \hline
	\end{tabular}
	\caption{\label{tab:table-name}Means and Standard Deviations of important features}
	\end{center} 
\end{table}

In terms of unique words, we found that RR used significantly more unique words, non-specific nouns and low imageability verbs than GHWB and DJT (see Table 3.3). Some of these differences are due to the length of the sample, particularly in the case of DJT where his average sample is almost half the sample of RR. It could also be said that this could be down to differences in linguistic abilities or speaking style \cite{Berisha2015, Le2011}. However, we can certainly see that as controls GWHB and DJT are comparative in relation to non-specific nouns and LI verbs.

@@

We then looked at the data from a longitudinal perspective as we are interested seeing whether we can track various language variables and their progress over time. We ran a number of Pearsons correlations with transcript index number as a time reference and the dependant variables (Table 3.4).  For our controls, we found them to be stable for the most part with the main highlights being a decrease in Adverb usage for DJT (R=-0.36, p=0.049) and a steady but not severe decline in a number of variables for GWHB, namely total word count, unique words, low imageability words and verb usage.
\par 
For RR, his decline is more marked and more widespread through his language use. We noticed an significant increase in adverb (R=0.41, p=0.004) and pronoun usage (R=0.65, p\textless0.001), as well as a slight usage increase in Non-specific nouns(R=0.30, p=0.03). There was a highly significant decrease in number of unique words (R=-0.56, p\textless0.001) and noun usage (R=-0.70, p\textless0.001). Also very significant decrease in adjective usage (R=-0.40, p=0.005) and a significant decrease in total word count (R=-0.31, p=0.03).

@@

Given the number of hypotheses being tested, it is necessary to apply some correction for family wise error rate and therefore a Hochberg step-up procedure was applied to the results of the analysis. Because features have been generated from multiple sources, there are some cases where there is significant overlap between the features.  Where possible the most appropriate feature has been selected, for example in the case of the raw count of pronouns and pronouns normalised by the length of the document, the latter has been selected. Where there is doubt as to the most appropriate feature, all features have been included.

@@

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time.
\par

A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample.  
\par 
Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Almor1999, Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
One of the most striking differences between this study and the original is when looking at Unique Words. Whilst Unique Words as a feature is significant, which was a finding in the original study, when we controlled via normalisation, we found that this was not significant.

>>>>

\usepackage{arxiv}
\usepackage{float}
\usepackage{graphicx}

@@

Alzheimer's Disease (AD) and other forms of dementia affect a significant proportion of the global older adult population and the impact on morbidity and mortality rates is considerable. Dementia and AD are currently the leading cause of death in England and Wales and the sixth leading cause of death in the United States (US). A recent report commissioned by the Alzheimer's Society in 2015 estimated that in the UK by 2015 there would be approximately 855,000 people rising to 1 million by 2021. This represents 1 in 79 of the general population rising to 1 in 14 of those aged 65 or over \cite{AlzheimersSociety2014}. Worldwide, there are 46 million people with a diagnosis of dementia globally and that number is expected to hit 131.5 million by 2050 \cite{Prince2015}. From a financial perspective, they estimated annual spend on dementia healthcare is £4.3 billion of which approximately £85 million is spent on diagnosis. The total financial burden of dementia (excluding the costs associated with early onset dementia) is £26.3 billion annually. Globally, this picture is a lot bleaker. Another report by Alzheimer's Disease International suggests that in 2015 there were  The report also states that the worldwide cost of AD in 2018 is estimated to be in the region of one trillon US dollars.

@@

\subsubsection{Linguistic Inquiry and Word Count(LIWC)}
The LIWC is a text processor which analysis words within a given a document and compares the words in the LIWC dictionary file. The LIWC dictionary file contains over 6000 words which are categorised into approximately 90 different types. For example the word 'cried' is part of five word categories: sadness, negative emotion, overall affect, verbs and past focus \cite{Pennebaker2015}. Each word in the document or set of documents is searched for in the dictionary file and if found, the appropriate type is incremented by 1. What is output is an analysis of word usage in reference to each category. This is particularly useful in analysing both structural language changes but also the content or themes of language for a given transcript or set of transcripts.
\par 
As with Berisha and Liss, we analysed the transcripts over time in order to determine if there was an underlying temporal trend that may indicate that different parts of language increase or decrease over time. In AD, cognitive abilities are said to deteriorate over time and therefore it may be useful to analyse temporal trends that indicate a potential deterioration of cognition. In order to get a accurate measure of deterioration over time, we time stamped each transcript in relation to number of days from the first transcript to the date of the transcript being analysed as the press conferences were not evenly spaced out during the presidencies.

\section{Results}\label{results}
One of the most important thing to note is the wide variety of samples between the three presidents and also the varying timescales. RR participated in 46 press conferences over eight years (an average of 5.75 a year) which is the fewest number of press conferences given by an American president during their term of office. GHWB participated in 136 press conferences over four years (an average of 34 a year) and DJT participated in 29 press conferences to date (an average of 19.3 per year). Equally, there are differences in the average number of words. RR produced an average of 3424 words per conference compared to 2608 by GHWB (unpaired t = 4.434, p\textless 0.001) and DJT at 1849 words (unpaired t = 6.524, p\textless 0.001).

\begin{table}[H]
	\begin{center}
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} |}
		\hline
		& RR & GHWB & DJT \\ \hline
		Total Words & 3423.91 (416.42) & 2607.72 (1210.38) & 1848.65 (1549.38) \\ \hline
		Unique Words & 894.13 (85.15) & 667.76 (218.67) & 481.82 (221.29) \\ \hline
		Mean Length of Utterance & 23.17 (1.402) & 18.71 (2.067) & 13.84 (1.619) \\ \hline
	\end{tabular}
	\caption{\label{tab:table-name}Means and Standard Deviations of general features for each set of transcripts}
	\end{center} 
\end{table}

In terms of more specific language differences between the presidents, we found that RR used significantly more unique words, non-specific nouns and low imageability verbs than GHWB and DJT (see Table 3.3). The mean length of utterance for RR was significantly greater than that of GHWB and DJT. Some of these differences are due to the length of the sample, particularly in the case of DJT where his average sample is almost half the sample of RR. It could also be said that this could be down to differences in linguistic abilities or speaking style \cite{Berisha2015, Le2011}. However, we can certainly see that as controls GWHB and DJT are comparative in relation to non-specific nouns and LI verbs.

@@

\subsection{Longitudinal Analysis}
We then looked at the data from a longitudinal perspective as we are interested seeing whether we can track various language variables and their progress over time. We ran a number of Pearsons correlations with transcript index number as a time reference and the dependant variables (Table 3.4).  For our controls, we found them to be stable for the most part with the main highlights being a decrease in Adverb usage for DJT (R=-0.36, p=0.049) and a steady but not severe decline in a number of variables for GWHB, namely total word count, unique words, low imageability words and verb usage.
\par 
For RR, his decline is more marked and more widespread through his language use. We noticed an significant increase in adverb (R=0.41, p=0.004) and pronoun usage (R=0.65, p\textless0.001), as well as a slight usage increase in Non-specific nouns(R=0.30, p=0.03). There was a highly significant decrease in number of unique words (R=-0.56, p\textless0.001) and noun usage (R=-0.70, p\textless0.001). Also very significant decrease in adjective usage (R=-0.40, p=0.005) and a significant decrease in total word count (R=-0.31, p=0.03).

@@

Given the number of hypotheses being tested, it is necessary to apply some correction for family wise error rate and therefore a Hochberg step-up procedure was applied to the results of the analysis. We also used the Benjamini-Yekutieli procedure for controlling for False Discovery rate and came up with a final list of important features for each president. For RR, this was 22 features, for GHWB this was 14 features and there were no significant features for DJT after applying these methods.  The features were generated from multiple sources and therefore there was some significant overlap between the features. In these cases, the normalised features were prefered to raw count features. Where there was doubt as to the most appropriate feature, all features have been included.

@@

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time.
\par 
A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample. In this paper, we controlled for transcript length by dividing any features that were raw counts by the total length of the transcript. Given the increased amounts of hypothesis testing, it became necessary to control for false discoveries. 
\par 
Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Almor1999, Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
One of the most striking differences between this study and the original is when looking at Unique Words. Whilst Unique Words as a feature is significant, which was a finding in the original study, when we controlled via normalisation, we found that this was not significant. 
\par
Another observation is that when conducting fitting a linear model to this data does not tell the whole story. For example, consider the following two plots. One shows a model fit to a linear  model, and the other shows a model fit using local polynomial regression model which aims to model the data more closely.

\begin{figure}[H]
\centering
\includegraphics[width=300px, height=300px]{RRppromLM.jpg}
\caption{RR Personal Pronouns - Linear Model}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=300px, height=300px]{RRppromLM.jpg}
\caption{RR Personal Pronouns - Local Regression Model}
\end{figure}
